Preventing a Nuclear ISIL: \\Recommendations for United States Policy

### 2015-05-12 draft.tex
Final version!!!

Where appropriate, expected countermeasures, such as shielding of radioactive sources, are discussed within the context of discerning a weak signal in a strong background. In circumstances where it is deemed essential for personnel to monitor the shipment of regular commerce or travel, including but not limited to commercial or private traffic by land, sea, or air, portal detectors and active interrogation is considered. Such locations are qualified for observation which are considered high risk or of strategic value in fielding a conventional response to attack from ISIL on the U.S., allies, or any sovereign territory for which instability is a matter of U.S. National Security. It is assumed that ISIL will prioritize counter-value over counterforce targets, since counter-value targeting is more disruptive and more effective in a campaign of terror.

\enquote{One of the most important lessons of the Cold War was that incessantly worrying about the low-probability / high-impact cases was a mistake.}  - Kenneth Pollack \cite{pollack2014unthinkable}

The ISIL nuclear threat is, by definition, a low-probability/high-impact case as this report has shown.  Developing nuclear weapons amidst crippling sanctions, airstrikes, and a dedicated opposition has never been done, despite several nation states attempting to do so.  Therefore, we choose to not repeat the mistakes of history and propose that no changes to U.S. policy be implemented to prevent ISIL from obtaining nuclear weapons.

\section{??????? ORDERED LIST OF RECOMMENDATIONS HERE ????????}

@@

\chapter{???????  Redo captions, to have true-ordered caption citation numbers    ?????????}

>>>>

This report addresses the possibility of the Islamic State of Iraq and the Levant (ISIL) becoming a nuclear threat to the United States and its allies by the acquisition of nuclear weapons, special nuclear material, or radiological dispersal devices. In accordance with President Obama's National Security Strategy, which states that \enquote{there is no greater threat to the American people than \ldots the pursuit of nuclear weapons by violent extremists \ldots}, this report informs a policy targeted at the prevention of such a calamity \cite{Obama2010}.

The unique history, ideology, and current geographic influence of ISIL are illustrated. ISIL is unlike any adversary the United States has had in the past, and therefore the need to understand and prevent any pathways to acquisition of nuclear material that could lead to wielding a nuclear weapon is imperative.

Policy recommendations for preventing a nuclear capable ISIL were developed by modeling the steps ISIL would have to take to procure a nuclear device, from the inception of the idea, to staging and detonation. Each recommendation is meant to target a specific chain of a  logic model developed in this report, with the ultimate goal of destroying all paths ISIL may have to a nuclear device.

\begin{easylist}[enumerate]
& Policies specifically targeting \textbf{indigenous development:}
&& Targeting of foundational capabilities	
% &&& Income 
% &&& People and expertise 
% &&& Infrastructure
&& IAEA monitoring 
% &&& Additional Protocol – push for more signatures 
% &&& International and Trafficking Database – push for increased cooperation
& Policies specifically targeting \textbf{proliferation through theft/purchase:}
&& IAEA monitoring 
% &&& International and Trafficking Database – push for increased cooperation
% &&& Additional Protocol – push for more signatures 
&& Global Threat Reduction Initiative (GTRI) 
% &&& Push for the cooperation of more nations
&& Strengthening of multilateral security agreements
&& Nuclear Forensics Attribution Act (NFAA)
% &&& Explicit declaratory policies, explicit deterrence 
&& United Nations Security Council Resolution (UNSCR) 1540 
% &&& Additional provisions and security expert group
&& Second Line of Defense (SLD)
% &&& Additional foreign ports
& Policies specifically targeting \textbf{delivery/staging:}
&& Military Force Protection Condition (FPCON)
% &&& Increased military base safety
&& Advanced Radiation Portal Monitor (RPM) Plan
% &&& Increased domestic port monitoring and advanced monitor development
&& Second Line of Defense (SLD)
% &&& Monitor additional foreign ports
\end{easylist}

% The risk of ISIL becoming a nuclear threat is a low-probability/high-impact case. Thus, the policy recommendations made in this report are targeted to reducing the already-low probability of nuclear acquisition as close to zero as possible. This report therefore also illustrates a dissenting opinion, offering a viewpoint that the resources allocated to the recommended policies could be disproportionate to the reality of the risk, and that current policies addressing the risk are adequate.

Therefore, the dissenting opinion recognizes that the marginal cost to implement the proposed policies outweigh the marginal benefit and recommends to not repeat the mistakes of history by implementing the proposed policies of this report. As Ken Pollack once said, \enquote{One of the most important lessons of the Cold War was that incessantly worrying about the low-probability / high-impact cases was a mistake.}  
Instead, it is proposed that the U.S. implements no changes to U.S. policy specifically to prevent the low probability event that is ISIL obtaining nuclear weapons. This would entail \enquote{staying the course} with President Obama's current policy of disrupting, degrading, and ultimately destroying ISIL as a necessary and sufficient condition to prevent nuclear weapon acquisition, while not losing focus on the real ISIL national security challenges \cite{WhiteHouse2014}.

% The ISIL nuclear threat is, by definition, a low-probability / high-impact case as this report has shown.  Developing nuclear weapons amidst crippling sanctions, airstrikes, and a dedicated opposition has never been done, despite several nation states attempting to do so.  Therefore, we choose to not repeat the mistakes of history and propose that no changes to U.S. policy be implemented to prevent ISIL from obtaining nuclear weapons.

@@

% \chapter{???????  Redo captions, to have true-ordered caption citation numbers    ?????????}
